---
title: "Geoprocessing in R"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
## Setup
```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(readr)
```

## Lesson 1: Accessing Spatial Data Files
- There are several ways we typically get spatial data into R:
  + Load spatial files we have on our machine or from remote source
  + Load spatial data that is part of an R package
  + Grab data using API (often making use of particular R packages)
  + Converting flat files with x,y data to spatial data
  + Geocoding data

### Vector Data
For reading and writing vector data in R, the primary package you'll use is:
 * `sf`for vector formats such as [ESRI Shapefiles](https://en.wikipedia.org/wiki/Shapefile),  [GeoJSON](https://en.wikipedia.org/wiki/GeoJSON), and [GPX](https://en.wikipedia.org/wiki/GPS_Exchange_Format) - both packages use **OGR**, which is a library under the [GDAL](https://en.wikipedia.org/wiki/GDAL) source tree,under the hood
 
 We can quickly discover supported I/O vector formats with `sf`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
library(sf)
print(paste0('There are ',st_drivers("vector") %>% nrow(), ' vector drivers available using st_read or read_sf'))
```
`sf` can be used to read numerous file types:

- Shapefiles
- Geodatabases
- Geopackages
- Geojson
- Spatial database files

#### Shapefiles
Typically working with vector GIS data we work with ESRI shapefiles or geodatabases - here we have an example of how one would read in a shapefile using `sf`:
```{r}
# download.file("ftp://ftp.gis.oregon.gov/adminbound/citylim_2017.zip","citylim_2017.zip")
# unzip("citylim_2017.zip", exdir = ".")
citylims <- st_read("citylim_2017.shp")
```

```{r}
options(scipen=3)
ggplot(citylims) + geom_sf() + ggtitle('Oregon City Limits') 
```

#### Geodatabases
```{r}
download.file("ftp://ftp.gis.oregon.gov/adminbound/OregonStateParks_20181010.zip", "OregonStateParks.zip")
unzip("OregonStateParks.zip", exdir = ".")
fgdb = "OregonStateParks_20181010.gdb"

# List all feature classes in a file geodatabase
st_layers(fgdb)
```

Read in the layer
```{r}
# Read the feature class
parks <- st_read(dsn=fgdb,layer="LO_PARKS")
ggplot(parks) + geom_sf()
```

#### Geopackages
Here we read in a geopackage loaded as part of the `sf` package.  We do a couple things here - in `st_layers`, we see there is just one layer, so we don't need to specify `layer` as a parameter in `read_sf`.  Second, we use pipes and dplyr syntax to read in our file - a nice background resource that delves into using the `tidyverse` is [R for Data Science](https://r4ds.had.co.nz/index.html).

```{r}
st_layers(system.file("gpkg/nc.gpkg", package="sf"))
nc <- system.file("gpkg/nc.gpkg", package="sf") %>% read_sf() # reads in
glimpse(nc)
write_sf(nc, 'nc.gpkg', 'nc', delete_layer = TRUE)
```

#### Open spatial data sources
There's a wealth of open spatial data accessible online now via static URLs or APIs - just a few examples include [Data.gov](https://catalog.data.gov/dataset?metadata_type=geospatial), NASA [SECAC Portal](http://sedac.ciesin.columbia.edu/), [Natural Earth](http://www.naturalearthdata.com/), [UNEP GEOdata](http://geodata.grid.unep.ch/), and countless others listed here at [Free GIS Data](https://freegisdata.rtwilson.com/)

Below is an example of pulling in US states using the `rnaturalearth` package - note that the default is to pull in data as `sp` objects and we coerce to `sf`.  Also take a look at the chained operation using dplyr.  Try changing the filter or a parameter in ggplot.
```{r naturalearth, , message=FALSE, warning=FALSE, error=FALSE}
library(rnaturalearth)
library(dplyr)
library(ggplot2)
states <- ne_states(country = 'United States of America')
states_sf <- st_as_sf(states)
states_sf %>%
  dplyr::filter(!name %in% c('Hawaii','Alaska') & !is.na(name)) %>%
  ggplot + geom_sf()
```

```{r}
states <- read_sf('https://eric.clst.org/assets/wiki/uploads/Stuff/gz_2010_us_040_00_500k.json')
states <- states[!states$NAME %in% c('Hawaii','Alaska'),]
plot(states$geometry)
```

#### Reading in subsets of vector data
Imagine we have really large datasets we need to work with.  We can subset the data as we read-in using SQL with `sf` as in this toy example using states data above.  Let's try to read in just PNW states from the states .json data.  See [st_read](https://r-spatial.github.io/sf/reference/st_read.html) for another example of this using the `sf` package `nc` data.
```{r}
library(stringr)
query_str <- str_c('SELECT * FROM "gz_2010_us_040_00_500k" WHERE NAME IN (\'',
                     'Oregon', '\', \'','Washington','\',\'','Idaho','\')')                     
states <- read_sf('https://eric.clst.org/assets/wiki/uploads/Stuff/gz_2010_us_040_00_500k.json',query = query_str)
glimpse(states)
plot(states$`_ogr_geometry_`)
```

#### Read in OpenStreetMap data
The [osmdata](https://github.com/ropensci/osmdata) package is facilitates accesses the OpenStreetMap (OSM) database in R and allows us to pull OSM data directly into R data structures and work with in R.

First we'll find available tags to get foot paths to plot
```{r osmdata1 , message=FALSE, warning=FALSE, error=FALSE}
library(osmdata)
library(mapview)
head(available_tags("highway")) # get rid of head when you run - just used to truncate output
```

```{r osmdata2 , message=FALSE, warning=FALSE, error=FALSE}
footway <- opq(bbox = "corvallis oregon") %>% 
  add_osm_feature(key = "highway", value = c("footway","cycleway","path", "path","pedestrian","track")) %>% 
  osmdata_sf()
footway <- footway$osm_lines
rstrnts <- opq(bbox = "corvallis oregon") %>% 
    add_osm_feature(key = "amenity", value = "restaurant") %>%
    osmdata_sf()
rstrnts <- rstrnts$osm_points
mapview(footway$geometry) + mapview(rstrnts)
```

#### Exercise
Take a minute and try pulling in data of your own for your own area and plotting using `osmdata`

#### Spatial subset
We can also perform a spatial subset on read-in.  We extract our geometry list-column as well-known text and pass it as a filter on read in.
```{r}
wkt = st_as_text(st_geometry(states[1,])) # here we grab just WA from our PNW states in previous step
# filter by bbox of first feature geometry selected above:
states <- read_sf('https://eric.clst.org/assets/wiki/uploads/Stuff/gz_2010_us_040_00_500k.json',wkt_filter = wkt)
plot(states$geometry)
# why are Oregon and Idaho still included?
```

#### Exercise
Based on examples of reading in different datasets and subsetting, try to find a dataset online to read in, or try a different subset operation on example datasets so far, or both.

## Raster Data
For reading and writing raster data in R, the primary packages you'll use are:
 * `raster` and / or `terra` for raster formats such as [GeoTIFF](https://en.wikipedia.org/wiki/GeoTIFF) or [ESRI or ASCII grid](https://en.wikipedia.org/wiki/Esri_grid) using [GDAL](https://en.wikipedia.org/wiki/GDAL) under the hood

As well as I/O raster formats via `sf`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
print(paste0('There are ',st_drivers(what='raster') %>% nrow(), ' raster drivers available'))
kable(head(st_drivers(what='raster'),n=5))
```